Scan-Centric, Frequency-Based Method for Characterizing Peaks from Direct Injection Fourier transform Mass Spectrometry Experiments

### 2022-04-11 doc/peakcharacterization_manuscript.Rmd
adding results for coefficients qcqa

@@

::: {custom-style="MDPI_3.1_text"}
Thermo-Fisher XCalibur was used to export peak lists from all samples used.
SMIRFE v 1.0 [@mitchellSmallMoleculeIsotope2019] running under Python 3.8 [@vanrossumPython3] was used for assignments.
Lipid classifications were generated by LipidClassifier v 1.0 [@mitchellDerivingLipidClassification2020].
All other calculations were performed in R v 4.1.0 [@rcoreteam_rlanguage_2020].
The targets package v 0.10.0 was used to control processing and aggregation of results [@landauTargetsPackageDynamic2021], and renv v 0.15.3 [@usheyRenvProjectEnvironments2022] to create a reproducible R package environment.
Plots were generated using ggplot2 v 3.3.5 [@wickhamGgplot2ElegantGraphics2016], patchwork v 1.1.1 [@pedersenPatchworkComposerPlots2020], ComplexHeatmap v 2.10.0 [@guComplexHeatmapsReveal2016a], ggridges v 0.5.3 [@wilkeGgridgesRidgelinePlots2021], and ggforce v 0.3.3 [@pedersenGgforceAcceleratingGgplot22021].
MSnbase v 2.20.4 [@gattoMSnbaseEfficientElegant2020; @gattoMSnbaseBioconductorPackage2012] provided facilities for reading in scan-level data, merging scans and calculating centroided peaks.
ICI-Kt correlation values among lipid samples were calculated using ICIKendallTau v 0.1.16 [@flightInformationContentInformedKendalltauCorrelation2022].
Outlier lipid samples were determined using visualizationQualityControl v 0.4.7 [@flightVisualizationQualityControlDevelopmentVisualization2021].
Specific data manipulation facilities were provided by dplyr v 1.0.8 [@wickhamDplyrGrammarData2022], tidyr v 1.2.0 [@wickhamTidyrTidyMessy2022], furrr v 0.2.3 [@vaughanFurrrApplyMapping2021a], 
This manuscript was generated from rmarkdown v 2.11 [@allaireRmarkdownDynamicDocuments2021; @xieMarkdownDefinitiveGuide2018; xieMarkdownCookbook2020].
:::

>>>>

::: {custom-style="MDPI_2.2_heading2"}
### Quality Control and Quality Analysis
:::

```{r load_increment_qcqa}
figure_count$increment("coefficients_qcqa")
figure_count$increment("frequency_mode_qcqa")
tar_load(coefficients_qcqa)
```

::: {custom-style="MDPI_3.1_text"}
Having a scan-centric workflow for generating the peak centroids means that we have opportunities to evaluate the scans as a whole, as well as the peaks across scans.
For example, we can mark peaks with unusually high frequency standard deviations (FSD) and mitigate the presence of high-peak-density artefacts (see **Mitigation of High Peak Density Artefacts**).
We also sometimes find that from initial peak region that has been marked as a single region, still results in two peaks being characterized in a single scan.
This results in that peak being removed from that scan entirely.

The conversion of m/z to frequency also provides opportunities to evaluate each scan and whether it should be kept for further processing.
One way is to examine the coefficients of the square-root term for each scan (see term **y** in Equation `r equation_count$label_text("mzfrequency")`) and look for outliers there.
Across the `r length(coefficients_qcqa$n_coef_out)` raw files that completed scan-centric processing, `r sum(coefficients_qcqa$n_coef_out > 0)` had at least one scan removed based on the square-root coefficient.
`r figure_count$label_text("coefficients_qcqa")`A shows a histogram of the number of scans removed based on being outliers of the overall distribution.
`r figure_count$label_text("coefficients_qcqa")`B shows what outlier scan coefficients look like in a single sample.
:::

::: {custom-style="MDPI_5.2_figure"}
```{r coefficients_qcqa}
coef_df1 = data.frame(N_out = coefficients_qcqa$n_coef_out)
coef_g1 = coef_df1 %>%
  ggplot(aes(x = N_out)) + 
  geom_histogram() +
  labs(x = "# of Scan Outliers", y = "# of Samples")

coef_g2 = coefficients_qcqa$example %>%
  ggplot(aes(x = sample, y = coefficient, color = outlier, group = sample)) +
  geom_sina(size = 2) +
  labs(x = "", y = "Square-Root Term") +
  theme(legend.position = c(0.1, 0.8))

(coef_g1 | coef_g2) + plot_annotation(tag_levels = "A")
```
:::

::: {custom-style="MDPI_5.1_figure_caption"}
`r figure_count$label_text("coefficients_qcqa")`.
**A**: Number of scan outliers for each sample in the NSCLC dataset.
**B**: Single example for sample 199Cpos of the scan square-root terms marked as outliers and removed from further consideration.
:::

One other check we implemented in the conversion to frequency is checking the range of frequency point-point differences allowed for creating the m/z to frequency model, and making sure that they are the same across scans.
The initial set of scans we allowed to use included **all** of the MS1 scans, including the precursor scans for the MS2 scans. 
The samples were acquired as a mix of MS1 only scans for the first 7.5 minutes, and then a mix of MS1 precursor scans followed by MS2 scans for the second 7.5 minutes.
In 20 samples, the MS1 precursor scans have a different mode in the frequency point-point difference than the primary MS1 scans, as shown in `r figure_count$label_text("frequency_mode_qcqa")`.

@@

::: {custom-style="MDPI_3.1_text"}
Thermo-Fisher XCalibur was used to export peak lists from all samples used.
SMIRFE v 1.0 [@mitchellSmallMoleculeIsotope2019] running under Python 3.8 [@vanrossumPython3] was used for assignments.
Lipid classifications were generated by LipidClassifier v 1.0 [@mitchellDerivingLipidClassification2020].
All other calculations were performed in R v 4.1.0 [@rcoreteam_rlanguage_2020].
The targets package v 0.10.0 was used to control processing and aggregation of results [@landauTargetsPackageDynamic2021], and renv v 0.15.3 [@usheyRenvProjectEnvironments2022] to create a reproducible R package environment.
Plots were generated using ggplot2 v 3.3.5 [@wickhamGgplot2ElegantGraphics2016], patchwork v 1.1.1 [@pedersenPatchworkComposerPlots2020], ComplexHeatmap v 2.10.0 [@guComplexHeatmapsReveal2016a], ggridges v 0.5.3 [@wilkeGgridgesRidgelinePlots2021], and ggforce v 0.3.3 [@pedersenGgforceAcceleratingGgplot22021].
MSnbase v 2.20.4 [@gattoMSnbaseEfficientElegant2020; @gattoMSnbaseBioconductorPackage2012] provided facilities for reading in scan-level data, merging scans and calculating centroided peaks.
ICI-Kt correlation values among lipid samples were calculated using ICIKendallTau v 0.1.16 [@flightInformationContentInformedKendalltauCorrelation2022].
Outlier lipid samples were determined using visualizationQualityControl v 0.4.7 [@flightVisualizationQualityControlDevelopmentVisualization2021].
Specific data manipulation facilities were provided by dplyr v 1.0.8 [@wickhamDplyrGrammarData2022], tidyr v 1.2.0 [@wickhamTidyrTidyMessy2022], furrr v 0.2.3 [@vaughanFurrrApplyMapping2021a], 
This manuscript was generated from rmarkdown v 2.11 [@allaireRmarkdownDynamicDocuments2021; @xieMarkdownDefinitiveGuide2018; @xieMarkdownCookbook2020].
:::
